---
title: "Benchmark TI Tools on Cancer Single Cell Data"
author: "Sehyun Oh"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 4
vignette: >
  % \VignetteIndexEntry{Statistical analysis output}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(cowplot)
library(ggplot2)
```

Combine tumor (196) and normal (355) cells  

# Load cell type information
```{r}
wd = "/data2/Understand-cancer-evolution-through-single-cell-expression-dynamics/presentations"
cellType = read.table(file.path(wd, "examples/GSM3348307_GSM3348308_barcode2celltype.tsv"),
                      sep = "\t", header = TRUE)
```

# Topology
## monocle3
```{r}
monocle = read.table(file.path(wd, "tmp/monocle3_barcode2pseudotime.tsv"), 
                     sep = "\t", header = TRUE)
colnames(monocle) = c("CellId", "monocle3")
```

## slingshot
```{r}
# slingshot_both <- readRDS(file.path(wd, "examples/pseudotime_combined_5and6.rds"))
slingshot = read.table(file.path(wd, "tmp/slingshot_both.csv"), sep = ",", header = TRUE)
colnames(slingshot)[1] = "CellId"
```

```{r}
cnv = read.csv(file.path(wd, "examples/examplesCNV_inference.tsv"), sep = "\t", header = TRUE)
cnv$CellId = gsub("-4", "", cnv$CellId)
cnv$CellId = gsub("-5", "", cnv$CellId)
```

## Plantir
```{r}
plantir = read.table(file.path(wd, "examples/palantir_barcode2pseudotime.csv"), sep = ",")
colnames(plantir) = c("CellId", "Plantir")
```

## PAGA
```{r}
paga = read.table(file.path(wd, "examples/paga_pdt.tsv"), sep = "\t", header = TRUE)
colnames(paga) = c("CellId", "paga")
paga$CellId = gsub("-4", "", paga$CellId)
paga$CellId = gsub("-5", "", paga$CellId)
```


# Pseudotime
We compared the pseudotime of each cell from different tools.

## slingshot vs. monocle3
```{r}
compare1 = dplyr::full_join(monocle, slingshot, by = "CellId")
compare1 = compare1[!is.na(compare1$slingPseudotime_1),]
# compare1 = compare1[sort(compare1$slingPseudotime_1),]
```

```{r echo=FALSE,fig.height=5, fig.width=6}
df = compare1
x = "slingPseudotime_1"
y = "monocle3"
cor = FALSE

# customLabs = labs(x = "SNP6 array (ABSOLUTE)", 
#                   y = "Tumor WES (PureCN)", 
#                   title = "A. TCGA-OV Purity") 

customLabs = labs(x = "Pseudotime_1 from slingshot",
                  y = "Pseudotime from monocle3")

source(file.path(wd, "tmp/figure_style.R"))
```

## Plantir vs. monocle3
```{r}
compare4 = dplyr::full_join(monocle, plantir, by = "CellId")
```

```{r echo=FALSE,fig.height=5, fig.width=6}
df = compare4
x = "Plantir"
y = "monocle3"
cor = FALSE

# customLabs = labs(x = "SNP6 array (ABSOLUTE)", 
#                   y = "Tumor WES (PureCN)", 
#                   title = "A. TCGA-OV Purity") 

customLabs = labs(x = "Pseudotime from Plantir",
                  y = "Pseudotime from monocle3")

source(file.path(wd, "tmp/figure_style.R"))
```

## Plantir vs. slingshot
```{r}
compare = dplyr::full_join(slingshot, plantir, by = "CellId")
```

```{r echo=FALSE,fig.height=5, fig.width=6}
df = compare
x = "Plantir"
y = "slingPseudotime_1"
cor = TRUE

customLabs = labs(x = "Pseudotime from Plantir",
                  y = "Pseudotime from slingshot")

source(file.path(wd, "tmp/figure_style.R"))
```

## monocle vs. PAGA
```{r}
compare = dplyr::full_join(paga, monocle, by = "CellId")
```

```{r echo=FALSE,fig.height=5, fig.width=6}
df = compare
x = "paga"
y = "monocle3"
cor = TRUE

customLabs = labs(x = "Pseudotime from PAGA",
                  y = "Pseudotime from monocle3")

source(file.path(wd, "tmp/figure_style.R"))
```






# CNV vs. Pseudotime
`CNV_inverence` represents the standard deviation of copy number changes in each cell.

## CNV vs. monocle3
```{r}
compare2 = dplyr::full_join(cnv, monocle, by = "CellId")
# compare2 = compare2[sort(compare2$monocle3),]
```

```{r echo=FALSE, fig.height=5, fig.width=6}
df = compare2
x = "monocle3"
y = "CNV_inference"
cor = FALSE

customLabs = labs(x = "monocle3",
                  y = "CNV inference")

source(file.path(wd, "tmp/figure_style.R"))
```

## CNV vs. slingshot
```{r}
compare3 = dplyr::full_join(cnv, slingshot, by = "CellId")
# compare = compare[!is.na(compare$slingPseudotime_1),]   # 439 cells
```

```{r fig.height=5, fig.width=6}
df = compare3
x = "slingPseudotime_1"
y = "CNV_inference"
cor = FALSE

customLabs = labs(x = "Pseudotime_1 from slingshot",
                  y = "CNV inference")

source(file.path(wd, "tmp/figure_style.R"))
```

## CNV vs. Plantir
```{r}
compare5 = dplyr::full_join(cnv, plantir, by = "CellId")
```

```{r fig.height=5, fig.width=6}
df = compare5
x = "Plantir"
y = "CNV_inference"
cor = FALSE

customLabs = labs(x = "Pseudotime from Plantir",
                  y = "CNV inference")

source(file.path(wd, "tmp/figure_style.R"))
```

## CNV vs. Plantir
```{r}
compare6 = dplyr::full_join(cnv, paga, by = "CellId")
```

```{r fig.height=5, fig.width=6}
df = compare6
x = "paga"
y = "CNV_inference"
cor = FALSE

customLabs = labs(x = "Pseudotime from PAGA",
                  y = "CNV inference")

source(file.path(wd, "tmp/figure_style.R"))
```